Møil: A web-based administration tool for mail servers backed by relational databases

### 2014-07-24 doc/paper/paper.tex
Added ERD.
Acronyms and text on data model criteria.

%\usepackage{german}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}

@@

\usepackage{biblatex}
%\usepackage[style=alphabetic]{biblatex}
\ExecuteBibliographyOptions{hyperref=true,language=ngerman,backref=true}
\addbibresource{bibliography.bib}

%\usepackage{lineno}
%\linenumbers

@@

	\section*{Abstract}
		This document compares existing administration user interface solutions
		for mail servers backed by relational databases and introduces yet
		another one, which is called Møil and is written in Ruby utilizing the
		Ruby on Rails\footnote{\url{http://rubyonrails.org}} framework. It
		brings the handy possibility of managing the database schema with
		migrations, a lot of beautiful, responsive CRUD and some other nice
		features (see page \pageref{moeil-features}). The code is available from
		GitHub\footnote{\url{https://github.com/nning/moeil}} licensed under
		AGPLv3 \cite{agpl}.

@@

		Database Management Systems (DBMS) for relational databases usually
		provide a query interface with the Structured Query Language (SQL)
		\cite{sql} for editing the data sets. It is widely supported by
		programming languages and equally widely spoken among programmers, which
		makes it valuable for integration of several services dealing with this
		mail system meta data. For example this can be used to integrate
		accounting solutions for charging purposes into mail systems.

@@

		Another reason for the usage of relational databases as mail system
		meta data storage is the increased structural clarity for
		administrators. In a typical SMTP server setup (without a relational
		database), the relevant data for domains, mailboxes and aliases is
		distributed to several files. The existence of a domain is stated
		independently of the existence of a mailbox of the same domain or the
		password of this mailbox. The data is not interconnected; a mailbox
		address is usually not checked against the list of domains for
		consistency. For bigger mail system setups, this meta data
		configuration inevitably gets unmaintainable.

@@

		Although an extensive evaluation of the existing administration user
		interfaces did not happen before implementing a custom solution, some
		of the programs were still sighted and some of the following criteria
		was still considered. This more systematical and detailed analysis
		shall gain an overview and provide inspiration for the development of
		Møil.

		\subsection*{Criteria}
			% Data model
			%	Structure (postfix.admin compatible?)
			%	Maintainability of schema
			%	Security (e.g. password hashing)
			% Responsiveness
			%	Terminal
			%	Desktop web browsers
			%	Mobile devices
			% Usability
			% Scalability (for user exposed functionality)
			% Features
			%	Basic domain, mailbox and alias management
			%	Password change for end users
			%	Privilege model
			%	Others
			%		Audit trail
			%		Vacation
			%		Spam frontend
			%		Sieve
			%		Webmail integration
			% Programming language
			% Containability (e.g. independence of Message Store file system)
			% Database independence
			% Security, robustness, tests

@@

			\subsubsection*{postfix.admin}
				\infotable{http://postfixadmin.sourceforge.net}{PHP}

@@

		\label{moeil-features}
		\subsection*{Features}

@@

		\subsection*{Data model}
			% Data model
			%	Structure / ERD
			%	Migrations
			%	sha512-crypt

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[ampersand]{easylist}
\usepackage[printonlyused,withpage]{acronym}

@@

\usepackage{biblatex}
\ExecuteBibliographyOptions{hyperref=true,language=ngerman,backref=true}
\addbibresource{bibliography.bib}

@@

	\section*{Abstract}
		This document compares existing administration user interface solutions
		for mail servers backed by relational databases and introduces yet
		another one, which is called \emph{Møil} and is written in Ruby
		utilizing the Ruby on Rails\footnote{\url{http://rubyonrails.org}}
		framework. It brings the handy possibility of managing the database
		schema with migrations, a lot of beautiful, responsive \ac{CRUD} and
		some other nice features (see page \pageref{sec:moeil:features}). The
		code is available from
		GitHub\footnote{\url{https://github.com/nning/moeil}} licensed under
		AGPLv3 \cite{agpl}.

@@

		\acp{DBMS} for relational databases usually provide a query interface
		with the \ac{SQL} \cite{sql} for editing the data sets. It is widely
		supported by programming languages and equally widely spoken among
		programmers, which makes it valuable for integration of several
		services dealing with this mail system meta data. For example this can
		be used to integrate accounting solutions for charging purposes into
		mail systems.

@@

		Another reason for the usage of relational databases as mail system
		meta data storage is the increased structural clarity for
		administrators. In a typical \acs{SMTP} \cite{smtp} server setup
		(without a relational database), the relevant data for domains,
		mailboxes and aliases is distributed to several files. The existence of
		a domain is stated independently of the existence of a mailbox of the
		same domain or the password of this mailbox. The data is not
		interconnected; a mailbox address is usually not checked against the
		list of domains for consistency. For bigger mail system setups, this
		meta data configuration inevitably gets unmaintainable.

@@

		Although an extensive evaluation of the existing administration user
		interfaces did not happen before implementing a custom solution, some
		of the programs were still sighted and some of the following criteria
		was still considered. This more systematical and detailed analysis
		shall gain an overview and provide inspiration for the development of
		\emph{Møil}.

		\subsection*{Criteria}
			Not only for the development of \emph{Møil}, some of the criteria
			arose from shortcomings of \emph{postfix.admin} (see page
			\pageref{sec:contestants:postfix.admin}), which is a quite widely
			deployed solution.

			\subsubsection*{Data model (1)}
				From a database engineering perspective, it is in general
				favourable to design a database with a \crit{normalized
				relational database schema}{1.1} \cite{dbnorm}. For a query to
				check the existence of a mailbox for example, a JOIN or nested
				query would be necessary, which has a very small performance
				impact. However, the schema normalization prevents data
				redundancies and inconsistencies. \emph{postfix.admin} does not
				have a normalized schema, so any solution which has one can not
				be \crit{compatible to \emph{postfix.admin}}{1.2}. A
				compatibility could be useful, if there is a big database
				already administered with \emph{postfix.admin}, which shall be
				migrated to another tool. Another possibility for this is an
				\crit{import of \emph{postfix.admin} data}{1.3}. When it comes
				to updates to the administration interface, \crit{means to
				handle database schema updates}{1.4} ease the update process
				(and therefore increase the probability of the actual
				conduction of an update). Last but not least, \crit{database
				independence}{1.5} allows a free choice of the \ac{DBMS} and
				integration with other software.

			\subsubsection*{Numbered index}
				\begin{easylist}
					& Data model
					&& Normalized schema
					&& \emph{postfix.admin} compatible
					&& \emph{postfix.admin} import
					&& Updatable (Migrations)
					&& Database independence

					& User interface
					&& Responsiveness
					&&&	\ac{CLI} support
					&&& Mobile devices
					&& Usability

					& Features
					&& Basic domain, mailbox and alias management
					&& Password changing for end users
					&& Privilege seperation
					&& Others
					&&& Audit trail
					&&& Vacation
					&&& Spam frontend
					&&& Sieve

					& Security \& robustness
					&& Password hashing
					&& Tests

					& Miscellaneous
					&& Scalability
					&& Programming language
					&& Containability (e.g. independence of Message Store file system)
				\end{easylist}

@@

			\label{sec:contestants:postfix.admin}
			\subsubsection*{postfix.admin}
				\infotable{http://postfixadmin.sourceforge.net}{PHP}

				As said, \emph{postfix.admin} seems to be a (negative) example
				not only for \emph{Møil}. Many solutions aim to overcome its
				shortcomings. \emph{postfix.admin} uses a non-normalized
				\cite{dbnorm} database schema, which could spare a JOIN in some
				database queries but is very prone to inconsistencies. There
				are no means of handling database schema updates; the schema is
				imported via a \ac{SQL} file on the initial installation. Also
				it creates an alias entry for every mailbox (even when the mail
				is not forwarded). This may have a cause, but is messy anyway.

@@

		\label{sec:moeil:features}
		\subsection*{Features}

@@

		\subsection*{Data model}
			% Data model
			%	Structure / ERD
			\includegraphics[width=\textwidth]{images/erd.pdf}
			%	Migrations
			%	sha512-crypt

@@

	\include{acronyms}
